Structural controllability of ecological networks

### 2018-06-22 paper/manuscript.Rmd
First draft of discussion

@@

Here, we apply methods from control theory to a particular ecological problem and show how it can be used to gain insight into the role of species in an ecological network. 
Specifically, we outline the approach using a set of ten pairs of uninvaded and invaded plant-pollinator communities. 
We use invaded communities because there is strong empirical evidence showing that invasive species play an important role shaping the abundances of other species, something which is particularly true in these ten networks [@lopezaraizamikel_impact_2007; @bartomeus_contrasting_2008-1]. 
This choice thus offers us an opportunity to explicitly contrast our theoretical observations with empirical evidence.
With these networks, we first ask whether there are differences between the controllability of invaded and uninvaded networks. 
This question is motivated by the difficulties commonly encountered with invasive-species eradication and ecosystem restoration [@woodford_confronting_2016].
<!--STOUFFER: I'd remove the grounded in.... stuff, but that's just me.  CAGUA: I don't like it either. I tried removing but then it feels like this question is coming out of the blue.-->
We then expand existing methods from control theory to effectively link the *controllability* of a network with the role of particular species.
We ask—from a control-theoretic perspective—whether there are key differences between species in the role they play at driving the population of other members of the community and identify the ecological factors related to these differences. 
This allows us to identify species that are critical to changes in the ecosystem state and show that they have a larger than expected impact on the structural stability of the community. 
Finally, we compare the proposed approach to current methods based on species centrality and show how these methods are indeed valuable but ultimately paint a limited picture in regards to the dynamic "keystoneness" of a species.

@@

>>>>

# Abstract

\clearpage

@@

Here, we apply methods from control theory to a particular ecological problem and show how it can be used to gain insight into the role of species in an ecological network. 
Specifically, we outline the approach using a set of ten pairs of uninvaded and invaded plant-pollinator communities. 
We use invaded communities because there is strong empirical evidence showing that invasive species play an important role shaping the abundances of other species, something which is particularly true in these ten networks [@lopezaraizamikel_impact_2007; @bartomeus_contrasting_2008-1]. 
This choice thus offers us an opportunity to explicitly contrast our theoretical observations with empirical evidence.
With these networks, we first ask whether there are differences between the controllability of invaded and uninvaded networks. 
<!--STOUFFER: I'd remove the grounded in.... stuff, but that's just me.  CAGUA: I don't like it either. I tried removing but then it feels like this question is coming out of the blue.-->
We then expand existing methods from control theory to effectively link the *controllability* of a network with the role of particular species.
We ask—from a control-theoretic perspective—whether there are key differences between species in the role they play at driving the population of other members of the community and identify the ecological factors related to these differences. 
This allows us to identify species that are critical to changes in the ecosystem state and show that they have a larger than expected impact on the structural stability of the community. 
Finally, we compare the proposed approach to current methods based on species centrality and show how these methods are indeed valuable but ultimately paint a limited picture in regards to the dynamic "keystoneness" of a species.

@@

Our main goal was to understand the role that species play at both modifying the abbundance of the species they interact with and the state of the community as a whole. 
To achieve that goal we applied *structural controllability*, a field at the intersection between control and complex theory that allow us to obtain two key pieces of information: the *controllability* of a network and a species *control capacity*.

Empirical observations indicate that steering the state of some communities—for example during ecosystem restoration or invasive species removal—can be a very difficult task [@woodford_confronting_2016]. 
Despite the large number of empirical examples, to date there is no thoretical basis that explains this difficutlies. 
Here, we show that the *controllability* of a network provides a way to quantify this intrinsic difficulty of controlling an ecological community and the differences among them. 
The controllability of a network is given by the relative size of the minimum driver node-set, $n_D$. 
In other words, controllability is related to the smallest proportion of species whose abundances need to be directly managed in order to achieve full control of the community (to be able to steer the community to an arbitrary ecological state). 
We found that in the pollination networks from our dataset the median $n_D$ equals `r round(median(c_m$n_D),2)`. 
This value is fairly high when compared to several other complex systems in which structural controllability has been applied. 
For instance, in most social, power transmission, internet, neuronal, and even metabolic networks $0.1 < n_D < 0.35$ and, at least withing the biological realm, only gene regulation networks appear to achieve similar levels of controllability [@liu_controllability_2011]. 
Although these results indicate that fully controlling ecological networks might currently be out of reach for all but the smallest communities [@motter_networkcontrology_2015], structural controllability might provide solid theoretical explanation to the many difficulties encountered in the management and restoration of natural communities.

Structural controllability might also offer novel means to compare ecological networks, particularly when exploring the relationship between network structure and stability. 
Even though all studied communities appear to be hard to control, we still found significant variation between them—the relative size of the minimum driver node set ranged between $n_D = `r round(min(c_m$n_D), 2)`$ and `r round(max(c_m$n_D), 2)`. 
Interestingly, we found that the controllability of a network is not related to their invasion status. 
Instead, we found that, despite our relatively small sample size, these differences in controllability may stem from differences in structure. 
Specifically, networks with a larger asymmetry between plant and pollinator species richness and with larger connectances were more likely to have higher values of $n_D$. 
In addition, we found that the $n_D$ of empirical networks was indistinguishable from that of network randomisations that roughly maintained the degree distribution, but was significantly different to those in which the patterns of dependence were broken. 
All together, these results suggest that network controllability is almost completely constrained by the patterns of species richness at each trophic guild and their degree distributions [@melian_complex_2002; @bluthgen_specialization_2007]. 
These two factors, in turn govern the asymmetric nature of mutual dependences-which constitute the foundations of structure and stability in mutualistic networks [@bascompte_asymmetric_2006; @memmott_tolerance_2004; @astegiano_robustness_2015].

Structural controllability not only can offer valuable insight at the community level, but, by measuring a species *control capacity*, here we show that it can also be used to better understand the role of individual species. 
The group of species that compose the minimum driver node-set $n_D$ is often not unique. 
This means that there are many species combinations with which it is possible to fully control a network
Each of these combinations represents a control configuration. 
Some species belong to these combinations more often that others. 
We call this property a species control capacity $\phi$. 
The higher a species control capacity, the higher the likelyhood that they would need to be directly managed to change (or maintain) the ecological state of their communities. 
A small fraction of species in each network (between `r round(pcrit_ran[1]*100)` and `r round(pcrit_ran[2]*100)`%) had a control capacity of $\phi = 1$. 
We call these *critical* species. 
Because critical species are driver in every single control configuration it is unlikely that any attempt to steer the community to a different state will be successful without directly managing the abbundances of these species. 
Interestingly, although the average control capacity of pollinators was higher than that of plants, the group of critical species was exclusively composed by plants.

Because of the presumed relationship between central and keystone species, we expected control capacity to be somewhat correlated with some metrics of centrality. 
However, despite the influence that critical species might have into the state of the community, a large control capacity is not necessarly accompained by a large centrality. 
Indeed,  the correlation coefficients between control capacity and five different centrality metrics was low and both specialist and generalist species could play an important role at changing the abbundances of other species.

Althogh control capacity was poorly predicted by centrality, we found that it was strongly related to a species contribution to nestedness and, to a smaller extent, the species strength. 
The species strength (the sum of visits that a species receivers or performs) quantifies how strongly a species interacts with its partners. 
Indirectly, it is also a reflection of the species' abundance, and therefore its relationship with control capacity can be explained by the fact that species with large populations are anticipated to have a larger effect in the community than less common species.

The strong relationship between control capacity and nestedness is more revealing in particular because we also found that critical species play a major role at enabling the coexistence of other species in their community. 
Previous studies suggest that nestedness is key to the persistence of pollination networks as it minimises species competition—and therefore promotes species coexistence—and confers robustness to species extinctions in the community. 
Using two completely different approaches, our results suggest that species that are more likely to determine the state of the community might also be more likely to contribute the most at "maintaining the organization and diversity of their ecological communities", one of the hallmarks of keystone species [@mills_keystone-species_1993].

Invasive species have been previously found to exacerbate the asymmetries in their communities [@aizen_invasive_2008; bartomeus_contrasting_2008-1; @henriksson_strong_2016]. 
Although this might be reflected on differences both at the community and the species level, we found that invasive plants are not inherently different to their native counterparts [@emer_species_2016; @stouffer_how_2014]. 
Invasive plants, just like any other mutualist in our data set, tended to attain a high control capacity proportional to the degree to which they contribute to nestedness and interact strongly with other species. 
Nevertheless 
Previous studies have found that supergeneralists, like invasive species, play a central role in their networks (Vilà et al. 2009; Palacio, Valderrama-Ardila & Kattan 2016). 
XXXOur results take this one step further and indicate that dependence strength, rather than generalism or other metrics of centrality, is the factor that best explains the cascading effects a species could trigger on its community.

The structural controllability approach, just like classic centrality metrics is sensitive to the uncertainity associated with the sampling of ecological interactions. 
Our sensitivity analyses suggest that structural controllability is robust to at least two sources of uncertainity: when interactions are only sampled incompletly and when multiple metrics could be used used to quantify the strength of associations between species.

Another common critique to structural controllability is that it is based on the premise that the system can be described using linear functional responses [@cowan_nodal_2012]. 
This is indeed a current limitation, as we know that ecological systems are often best described using different nodal dynamics. 
However, we are not discouraged by its application for three main reasons. 
First, it has been shown that when the parameters that describe the mechanisms of self-regulation are not independent from each other, as is ofthen the case in ecological networks, the principles of structuctural controllability can still be applied [@zhao_intrinsic_2015]. 
Second, although non-linear dynamics render structural controllability unsufficient to assure full control an ecological system, it is still a necessary step to achieve that goal. 
Finally, by accounting for network dynamics (even if in a simple way) it incorporates more ecological realism than classical centrality metrics and hence is more likely to provide useful information even in the extreme scenario in which the state of a community depends almost exclusively on nodal dynamics rather than on the topology of their interactions.

From an applied ecology perspective, @paine_note_1969 showed nearly 50 years ago that one species can dominate the state of its community. 
Structural controllability suggests that this situation might be the exception rather than the rule, at least when were interested in *managing* the state of 
community and not just on inducing a change. 
Our results also provide support to the idea that management approaches focused on ecosystem processes and interactions might be more effective than traditional single-species approaches [@harvey_bridging_2017]. 
A `r round(sum(crit$n_crit > 1)/nrow(crit)*100)`% of our communities contained more than one critical species. 
Approaches identify keystone species and ultimately prioritise species for conservation are still useful at predicting the consequences of single species removals or reintroductions [@pires_friendship_2017; @devoto_understanding_2012], but effective management and restoration of ecological communities is likely to require more than a ranking of species based on their traits or centrality.

<!-- Biggest lesson is that changing the state of the community in a requires concerted action by several species, not just changes in the abbundance of one species.  -->

<!-- Control theory is flexible and awknoledges that there are multiple ways to achieve changes in the state of the community.  -->
<!-- Managing the most central species might not be suficient or even required to achieve that goal.  -->

<!-- Even if full control is not possible or not desired we learnt that there are critical species, those that no matter what the management strategy will need to be managed if one wants effective control of the populations in the ecological community.  -->

<!-- Centrality species concept is limited because we onl -->

<!-- We propose structural controllbility as a metric that can be used to compare ecological systems both in a theoretical and applied context. -->
